Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2025-06-27 ms/ms.tex
read through on paper

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, \textgreater10 km\textsuperscript{2} sites.
We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
\end{abstract}

@@

Peat soils are a terrestrial carbon stock of global importance.
They store 450-650 Gt of carbon, or about 30 \% of global soil carbon, despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem \citep{temminkRecoveringWetlandBiogeomorphic2022}.
This makes them crucial to climate change mitigation.
Intact peatlands sequester carbon and overall produce a negative temperature forcing \citep{joostenRolePeatlandsClimate2016}.
When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.

One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth.
Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}, so deep peats comprise a large volume.
Their depth results from the accumulation of organic matter over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.

Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
However, other land uses also displace peatlands.
In Norway -- where flat lowland is relatively scarce but upward of 9 \% of the land area is peatland \citep{brynLandCoverNorway2018, bakkestuenDelineationWetlandAreas2023} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
Since then, a larger proportion of peatland loss is driven by construction \citep{flagetMyrMellomLokal2024}.

The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.

Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost.
However, surveying large areas is impractical when depth varies widely over short distances (e.g.~10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
A complementary approach from soil science is digital soil mapping (DSM).
DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model.
This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying statistical patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.

@@

Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary.
One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024, pohjankukkaDigitalMappingPeat2025}.
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.

@@

Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth.
Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.

The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area.
Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.

The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: \emph{shallow} (\textless1 m), or \emph{deep} (\textgreater1 m).
These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
The number of measurements per unit area was not standardized but probably low.

The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set.
With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.

Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway.
Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.

@@

At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig. \ref{fig:map-sites}a).
In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project.
The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is \(10.8^\circ\) (IQR \(4.6-15.1^\circ\)).

At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:map-sites}b).
In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is \(13.0^\circ\) (IQR \(4.7-18.3^\circ\)).

@@

Although radiometric data must be in units of counts per second to model attenuation directly \citep{olearyDigitalSoilMapping2022}, we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.

@@

We used the \emph{terra} package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2, Morphometry library,][]{conradSystemAutomatedGeoscientific2015} with default parameters (initial slope threshold = 16 \%, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).

@@

The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g.~stream, lake, or sea) that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
We calculated unitless slope from the 1 m DTM using the Whitebox software.
We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived Depth-to-Water using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

The places (10 m raster cells) we chose to measure peat depth were sampled from mire areas in the AR5 dataset, and optimized for training a Random Forest (RF) model of peat depth \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
We drew 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet.
A complete description of our sample selection is in Appendix \ref{sec:sample-selection}.

We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.

We also extracted a set of existing depth measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
We georeferenced the map and digitized the borehole locations and depths.

We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.

@@

We used Random Forests (RF) to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally predictive model aligns with the aim of this study.

>>>>

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, \textgreater10 km\textsuperscript{2} sites.
We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Radiometric variables showed inconsistent and weak predictive value -- improving performance at one site while degrading it at the other.
Our remote sensing models had better accuracy than the national map of peat depth, even when we calibrated the national map to the same depth data.
Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
Based on these findings, we conclude that digital soil mapping can improve the existing, national map of peat depth in Norway, but detailed local maps are best made from tailored field measurements.
Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
\end{abstract}

@@

Peat soils are a terrestrial carbon stock of global importance.
They store 450-650 Gt of carbon, or about 30 \% of global soil carbon, despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem \citep{temminkRecoveringWetlandBiogeomorphic2022}.
This makes them crucial to climate change mitigation.
Intact peatlands sequester carbon and overall produce a negative temperature forcing \citep{joostenRolePeatlandsClimate2016}.
When disturbed -- often by conversion to another land use -- they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.

One of the keys to the areal density of peatland carbon stocks (and variation therein) lies in the third dimension: their depth.
Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}.
Their depth depends on the rate and duration of organic matter accumulation, potentially over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.

Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils, but other land uses also displace peatlands.
In Norway -- where flat lowland is relatively scarce but upward of 9 \% of the land area is peatland \citep{brynLandCoverNorway2018, bakkestuenDelineationWetlandAreas2023} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
Since then, a larger proportion of peatland loss is driven by construction \citep{flagetMyrMellomLokal2024}.

Variation in peat depth and its spatial distribution is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
Maps are crucial because, unlike spatially aggregated estimates, they link high-level targets to specific management decisions \citep{oecdOECDEnvironmentalPerformance2022}.
Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.

Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost.
However, surveying large areas is impractical when depth varies widely over short distances (e.g., 10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
A complementary approach from soil science is digital soil mapping (DSM).
DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model.
This approach has grown in importance with the availability of remotely sensed data and the advancement of machine learning methods \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.

@@

Studies have consistently shown relationships between peat depth and topography, though the specific patterns vary.
One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.,][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.,][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth \citep[e.g.,][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024, pohjankukkaDigitalMappingPeat2025}.
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.

@@

Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
Radiometric data are increasingly available over large areas \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}, and -- insofar as they are predictive of peat depth -- this presents an opportunity for DSM.

The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area.
Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
These mesotopes have fundamentally different geomorphology, which suggests that relationships between peat depth and terrain or radiometric predictors may vary by landscape.

The only systematic mapping of peat depth at the national scale in Norway comes from surveys meant to identify arable land.
After scattered efforts in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: \emph{shallow} (\textless1 m), or \emph{deep} (\textgreater1 m).
These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
The number of measurements per unit area was not standardized but probably low.

The push for nature-based climate solutions motivates for extensive mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps available before detailed investigation increases the option space for spatial planning and land management.
For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work before road construction, but by that time, the route of the road is already set.
With a landscape-scale peat depth map, planners could better compare the climate impact of different routes.
To make landscape-scale maps possible, more studies need to investigate the relationships between peat depth and topographic or radiometric variables, to determine how strong and consistent these are.

Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view toward revising regional and national maps in Norway.
Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.

@@

At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig. \ref{fig:map-sites}a).
In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project.
The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
Its mean elevation is 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is \(10.8^\circ\) (IQR \(4.6-15.1^\circ\)).

At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:map-sites}b).
In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
This study area is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is \(13.0^\circ\) (IQR \(4.7-18.3^\circ\)).

@@

Although radiometric data must be in units of counts per second to model attenuation directly \citep{olearyDigitalSoilMapping2022}, we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
The monotonic transformation between counts per second and concentration has no effect on the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.

@@

We used the \emph{terra} R package (v1.8) to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2, Morphometry library,][]{conradSystemAutomatedGeoscientific2015} with default parameters (initial slope threshold = 16 \%, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).

@@

The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g., stream, lake, or sea) that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
We calculated unitless slope from the 1 m DTM using Whitebox.
We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived Depth-to-Water using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

The places (10 m raster cells) we chose to measure peat depth were sampled from mire areas in the AR5 dataset, and optimized for training a Random Forest (RF) model of peat depth \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
Although we implemented the sampling design differently at Skrimfjella (in 2020) than at Ørskogfjellet (in 2023), the overall approach was the same.
We chose 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet as our designed samples.
A complete description of our sample selection is in Appendix \ref{sec:sample-selection}.

@@

We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
We made at least three point measurements of peat depth within each of the raster cells in the designed samples.
At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
That is: peat depth was always measured by probing in our designed samples, and we also used GPR in a set of cells that partly overlapped with the designed samples.
We chose this combined approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.

At Ørskogfjellet we were also able to use two existing sets of peat depth measurements in addition to our own.
We extracted a set of 44 borehole depths (in decimeters) across a 9 ha peatland area, from a paper map made by the Norwegian Soil and Mire Company in 1984.
We also used a set of GPR-based depth measurements, commissioned and provided to us by the Norwegian Public Roads Administration.
These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
We used a total of 403440 interpreted and calibrated traces along 7.4 km of transects from this work -- discarding data where multiple depths were interpreted for the same location.

All point measurements described above were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

We used Random Forests (RF) to predict peat depth at both sites.
RF is an ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.

RF by itself is not a spatial model, and it will only predict spatial structure in the peat depth to the degree that spatial structure is captured by predictors.
We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that adds to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
The spatial component in regression kriging often improves map accuracy \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF was justified at both sites.
